---
title: "Clean Raw Data"
---

# Load raw data, clean it, and generate survey_wide_with_outcome for analysis

```{r}
library(assertr)
library(tidyverse)
```

# Define utilty functions

```{r}
# Adds a suffix to every variable name in the data set
add_suffix <- function(df, suffix){
  names(df) <- paste0(names(df), "_", suffix)
  df
}

# Replace a single value in the target column of a data frame. Target specified with filter column.
jz.replace_one_val <- function(df, col_filter, val_filter, col_target, val_target, dbg = FALSE) {
  r <- which(df[col_filter] == val_filter)
  stopifnot(length(r) == 1)
  if (dbg) 
    print(paste0("Replacing ", df[r, col_target], " with ", val_target))
  df[r, col_target] <- val_target
  df
}

# Replace a single value in the target column of a data frame only if it is NA. Target specified with filter column.
jz.replace_one_na_val <- function(df, col_filter, val_filter, col_target, val_target, dbg = FALSE) {
  r <- which((df[col_filter] == val_filter) & is.na(df[col_target]))
  stopifnot(length(r) == 1)
  if (dbg) 
    print(paste0("Replacing ", df[r, col_target], " with ", val_target))
  df[r, col_target] <- val_target
  df
}

# Replace a single value in the target column of a data frame only if it is zero. Target specified with filter column.
jz.replace_one_zero_val <- function(df, col_filter, val_filter, col_target, val_target, dbg = FALSE) {
  r <- which((df[col_filter] == val_filter) & (df[col_target] == 0))
  stopifnot(length(r) == 1)
  if (dbg) 
    print(paste0("Replacing ", df[r, col_target], " with ", val_target))
  df[r, col_target] <- val_target
  df
}

```

# Load raw data

```{r}
# Load the raw data
prin_survey_raw <- read_csv("data_raw/PrinSurvImpact.csv")
school_outcome_raw <- readxl::read_excel("data_raw/SchoolDV.xlsx")

# Verify that the data matches expectations
prin_survey_raw <- prin_survey_raw %>%
  verify(ncol(.) == 416) %>%
  verify(nrow(.) == 400)

school_outcome_raw <- school_outcome_raw %>%
  verify(length(names(.)) == 8) %>%
  verify(length(Treatment_OBS) == 67)
```

# Prune unnecessary variables

```{r}
prin_survey <- prin_survey_raw %>% 
  dplyr::select(-contains("zscore")) %>% 
  dplyr::select(intv_status:d5b_4)

school_outcome <- school_outcome_raw %>% 
  dplyr::select(-`In Survey?`) %>% 
  add_suffix("outcome")

```

# Fix up missing school pairings

```{r}
# Helper function to replace Treatment_OBS of 0 in survey
replace_zero_obs <- function(df, school_name, new_obs) {
  r <- which((df$SchoolName == school_name) & (df$Treatment_OBS == 0))
  stopifnot(length(r) == 1)
  df$Treatment_OBS[r] <- new_obs
  df
}

# Helper function to replace Treatment_OBS of NA in outcome
replace_na_obs <- function(df, school_name, new_obs) {
  r <- which((df$School_outcome == school_name) & is.na(df$Treatment_OBS_outcome))
  stopifnot(length(r) == 1)
  df$Treatment_OBS_outcome[r] <- new_obs
  df
}

# Survey fixups
prin_survey <- prin_survey %>%
  replace_zero_obs("BRIARMEADOW MIDDLE SCHOOL", 252) %>%  # YES Prep "LEE"
  replace_zero_obs("WOODSON MIDDLE", 230) %>%             # YES Prep "SOUTHWEST"
  replace_zero_obs("T H ROGERS SEC", 229)                 # YES Prep "NORTH CENTRAL"

# Outcome fixups
school_outcome <- school_outcome %>%
  replace_na_obs("Houston HSA Impact", 232) %>%           # Catch T H ROGERS
  replace_na_obs("WILL Academy Impact", 325) %>%          # Catch Stuart-Hobson Middle School
  replace_na_obs("Triumph Charter Academy", 78) %>%       # Catch SAN FERNANDO MIDDLE
  replace_na_obs("Dr. James L. Burch Intermediate", 203)  # Catch Nimitz Middle School

```

# Filter to CMOs with data

```{r}
# filter to the 19 for which we have outcome data
treat_cmo_good <- c(2, 4, 7, 8, 14, 16, 18, 19, 21, 22, 23, 24, 25, 26, 31, 32, 35, 39, 41)

prin_survey <- prin_survey %>% 
  filter(Treatment_CMOID %in% treat_cmo_good)

```

# Assign fake Treatment_OBS values starting at 500 to outcomes with missing district school pairings

```{r}
stopifnot(nrow(filter(school_outcome, is.na(Treatment_OBS_outcome))) == 18) # used to be 21 before fixups

school_outcome <- school_outcome %>%
  arrange(!is.na(Treatment_OBS_outcome), Treatment_OBS_outcome) %>%
  mutate(fake_OBS = 500:(499+nrow(school_outcome))) %>%
  mutate(Treatment_OBS_outcome = ifelse(is.na(Treatment_OBS_outcome), fake_OBS, Treatment_OBS_outcome)) %>%
  mutate(fake_OBS = NULL) %>%
  arrange(CMO_outcome)

stopifnot(nrow(filter(school_outcome, is.na(Treatment_OBS_outcome))) == 0)

```

# Create prin_survey_wide
NB: This removes the special Treatment_OBS == 0

```{r}
# spread wide
values <- sort(unique(prin_survey$Treatment_OBS))
values <- values[-1]

take <- function(t_obs){
  a <- prin_survey %>% 
  dplyr::select(Treatment_OBS, Admin, everything()) %>% 
  filter(Treatment_OBS == t_obs, Admin == "district") %>% 
  add_suffix("district")

  if(nrow(a) == 0){
    a[1, ] <- rep(NA, ncol(a))
  }

  b <- prin_survey %>% 
    dplyr::select(Treatment_OBS, Admin, everything()) %>% 
    filter(Treatment_OBS == t_obs, Admin == "CMO") %>% 
    add_suffix("CMO")

  if(nrow(b) == 0){
    b[1, ] <- rep(NA, ncol(b))
  }
  
  cbind(a, b)
}

# execute
prin_survey_wide <- values %>% map_df(take) %>% as_tibble()

```

# Add outcome data to prin_survey_wide to produce our final survey variable (formerly survey_wide_with_outcome)

```{r}
# Create fake observations in prin_survey_wide for the umatched outcomes so we can join the data sets
unmatched_OBS <- filter(school_outcome, Treatment_OBS_outcome >= 500)$Treatment_OBS_outcome

tmp <- prin_survey_wide %>% 
  mutate(Treatment_OBS = pmin(prin_survey_wide$Treatment_OBS_CMO, prin_survey_wide$Treatment_OBS_district, na.rm = TRUE)) %>% 
  add_row(Treatment_OBS = unmatched_OBS) %>%
  left_join(school_outcome %>% filter(!is.na(Treatment_OBS_outcome)), by = c("Treatment_OBS" = "Treatment_OBS_outcome")) %>% 
  dplyr::select(contains("Treatment_OBS"), contains("outcome"), everything()) 

survey <- tmp %>% 
  rename(math_outcome = `2YRMath_outcome`, reading_outcome = `2YReading_outcome`) %>% 
  filter(!is.na(math_outcome) | !is.na(reading_outcome)) 

```

# Fixup CMOID 

```{r}
# List of correct (Treatment_OBS, CMOID) pairs for Treatment_CMOID_CMO
fixup_CMOIDs <- list(
  c(78, 31),
  c(203, 35),
  c(232, 18),
  c(325, 25),
  c(500, 2),
  c(501, 2),
  c(502, 4),
  c(503, 8),
  c(504, 100),
  c(505, 100),
  c(506, 100),
  c(507, 100),
  c(508, 25),
  c(509, 101),
  c(510, 101),
  c(511, 101),
  c(512, 101),
  c(513, 35),
  c(514, 39),
  c(515, 102),
  c(516, 102),
  c(517, 102)
)

tmp <- t(simplify2array(fixup_CMOIDs))
for (i in 1:nrow(tmp)) {
  survey <- jz.replace_one_val(survey, "Treatment_OBS", tmp[i, 1], "Treatment_CMOID_CMO", tmp[i, 2])
}

```

# Fixup DistrictName

```{r}
# List all district names
tmp <- survey[c(which(names(survey) == "Treatment_OBS"), grep("DistrictName", names(survey)))]

# List all different district names
tmp2 <- subset(tmp, !is.na(DistrictName_CMO) & !is.na(DistrictName_district) & (DistrictName_CMO != DistrictName_district))
```

# Fixup incorrect states

```{r}
# List of correct (Treatment_OBS, state) pairs for CMO state
fixup_states <- list(
  c(78, "CA"),  # NA
  c(195, "TX"), # not PA
  c(203, "TX"), # NA
  c(232, "TX"), # NA
  c(254, "NY"), # not TX
  c(284, "AZ"), # not CA
  c(294, "NY"), # not TX
  c(299, "AZ"), # not CA
  c(309, "IL"), # not IN
  c(311, "TX"), # not NJ
  c(319, "AZ"), # not TX
  c(325, "DC"), # NA
  c(331, "TX"), # not DC
  c(500, "NY"), # NA
  c(501, "NY"), # NA
  c(502, "CA"), # NA
  c(503, "CA"), # NA
  c(504, "OH"), # NA
  c(505, "OH"), # NA
  c(506, "OH"), # NA
  c(507, "OH"), # NA
  c(508, "DC"), # NA
  c(509, "NY"), # NA
  c(510, "NY"), # NA
  c(511, "NY"), # NA
  c(512, "NY"), # NA
  c(513, "TX"), # NA
  c(514, "NY"), # NA
  c(515, "TX"), # NA
  c(516, "TX"), # NA
  c(517, "TX") # NA
  )

tmp <- t(simplify2array(fixup_states))
for (i in 1:nrow(tmp)) {
  survey <- jz.replace_one_val(survey, "Treatment_OBS", tmp[i, 1], "state_CMO", tmp[i, 2])
}

```

# Add a KIPP variable

```{r}
survey <- survey %>% 
  mutate(KIPP = ifelse(str_detect(str_to_lower(CMO_outcome), "kipp"), 1, 0))
stopifnot(sum(survey$KIPP) == 13)

```

# Compare the generated survey_wide_with_outcome with the existing one so we know if something changed

```{r}
# Save the current survey_wide so the column types match when comparing with the old one
survey %>% write_csv("data_created/survey_wide_with_outcome_tmp.csv")

old_sv <- read_csv("data_created/survey_wide_with_outcome.csv")
new_sv <- read_csv("data_created/survey_wide_with_outcome_tmp.csv")
all_equal(old_sv, new_sv)

```

# Save the survey_wide_with_outcome results to a csv file for subsequent analysis
NB: This code chunk will not run by default so we are intentional when we overwrite the file
NB: Remember to archive each version of survey_wide_with_outcome as we make changes

```{r, eval=FALSE}
survey %>% write_csv("data_created/survey_wide_with_outcome.csv")

```

