---
title: "DIF Example"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Takeways

We'll use a simulation to show how to identify DIF and what can go wrong. In particular, we'll see that 

1. If you anchor on an item without DIF, you recover everything well

2. If you accidentally anchor on an item with DIF, that difference gets incorrectly soaked up into the difference in group means

3. If you don't anchor on anything at all, the model isn't identified and you get weird results

## Simulation

Imagine two groups of students (old and young). 

First, simulate the 250k old students who have ability N(0, 1) and take a test where all items are Rasch and have item easiness of 0.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(mirt)

n_students_each_group <- 250000

theta_old <- rnorm(n_students_each_group, 0, 1)

data_old <- 
   simdata(
      a = c(1, 1, 1, 1, 1), # setting these all to 1 make these Rasch
      d = c(0, 0, 0, 0, 0),
      itemtype = "2PL",
      Theta = matrix(theta_old)
   )
```

Next simulate the 250k young students who have ability N(-1, 1) and take the same test with all Rasch items but for some reason the 5th item has easiness -1 instead of 0 so it is more difficult for these young students:

```{r}
theta_young <- rnorm(n_students_each_group, -1, 1)

data_young <- 
   simdata(
      a = c(1, 1, 1, 1, 1),
      d = c(0, 0, 0, 0, -1), # item 5 biases against young students!
      itemtype = "2PL",
      Theta = matrix(theta_young)
   )

# combine old and young data
data_combined <- rbind(data_old, data_young)

# and keep track of which rows are old vs. young
old_or_young <- rep(c("old", "young"), each = n_students_each_group)
```

To summarize, here are the data generating parameters:

```{r}
tibble(
   what = c(paste0("Item ", 1:5, " Easiness"), "Group Mean"),
   old = c(0, 0, 0, 0, 0, 0),
   young = c(0, 0, 0, 0, -1, -1)
) %>% 
   mutate(difference = young - old) %>% 
   kableExtra::kable()
```

We're going to fit a few multiple group models and we'll want to quickly summarize parameters so let's write a helper function:

```{r}
summarize_model <- function(model){
   coef_young <- coef(model, simplify = TRUE)$young
   coef_old <- coef(model, simplify = TRUE)$old
   
   tibble(
      what = c(paste0("Item ", 1:5, " Easiness"), "Group Mean"),
      old = c(coef_old$items[ , "d"], coef_old$means),
      young = c(coef_young$items[ , "d"], coef_young$means)
   ) %>% 
      mutate(difference = young - old) %>% 
      mutate_if(is.numeric, round, 3)
}
```

## 1: Choose a good anchor

Now we can fit multiple group models! 

To start, if we anchor on one or all of the items that don't have DIF (e.g. just item 1 in the code below) we happily recover the data generating parameters:

```{r}
multipleGroup(
   data_combined, 
   1, 
   itemtype = "Rasch",
   group = old_or_young, 
   # this notation says freely estimate the group means but
   # constrain item 1 to be invariant
   invariance = c("free_means", "Item_1"),
   verbose = FALSE
) %>% 
   summarize_model() %>% 
   kableExtra::kable()
```

## 2a: Constrain everything

If we accidentlly constrain all items (despite item 5 is being harder for young students), then that shows up in a couple of ways:

- lower estimate of the young group mean than we should have
- item 5 easiness is somewhere between the old value of 0 and the young value of -1

```{r}
multipleGroup(
   data_combined, 
   1, 
   itemtype = "Rasch",
   group = old_or_young, 
   # freely estimate group means but constrain all items invariant
   invariance = c("free_means", "intercepts"),
   verbose = FALSE
) %>% 
   summarize_model() %>% 
   kableExtra::kable()
```

## 2b: Choose a bad anchor

If we accidentally anchor item  5 which is variant then the difference in performance on item 5 gets added to the difference in group means:

```{r}
multipleGroup(
   data_combined, 
   1, 
   itemtype = "Rasch",
   group = old_or_young, 
   # this notation says freely estimate the group means but
   # constrain item 5 to be invariant
   invariance = c("free_means", "Item_5"),
   verbose = FALSE
) %>% 
   summarize_model() %>% 
   kableExtra::kable()
```

## 3: Don't anchor on anything

If we don't anchor on anything, then the model isn't identified and we get weird estimates. It's worth noticing that the sum of the difference in item 5 easiness and difference in group means still sums to -2 but it's spread out in an arbitrary way:

```{r}
multipleGroup(
   data_combined, 
   1, 
   itemtype = "Rasch",
   group = old_or_young, 
   # freely estimate group means (and don't constrain anything invariant)
   invariance = c("free_means"),
   verbose = FALSE
) %>% 
   summarize_model() %>% 
   kableExtra::kable()
```
