---
title: "easy_sim"
output: 
   github_document:
      toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Get started

```{r}
library(tidyverse)
set.seed(1)
```

# Part 1: Write a function

First write a function that simulates our data

```{r}
sim_tacos_and_points <- function(n_rows, mean_tacos, sd_tacos, sd_points, mean_points, ...){
   tibble(
      tacos_eaten = rnorm(n_rows, mean_tacos, sd_tacos),
      points_scored = rnorm(n_rows, mean_points, sd_points)
   )
}
```

Example using the function

```{r}
sim_tacos_and_points(
   n_rows = 10,
   mean_tacos = 5,
   sd_tacos = 3,
   mean_points = 10,
   sd_points = 3
)
```

# Part 2: Simulate data

Make a data frame of all parameter combinations we're interested in

```{r}
parameters <- 
   crossing(
      n_rows = 100,
      mean_tacos = c(5, 10),
      sd_tacos = c(1, 10),
      mean_points = c(10),
      sd_points = c(5)
   ) %>% print()
```

Row-wise simulate data and store in a column of dataframes.

```{r}
parameters <- 
   parameters %>% 
   mutate(data = pmap(., sim_tacos_and_points)) %>% 
   print()
```

Examples of a few things we can do with this

0) Extract just one of the datasets

```{r}
# based on row number
parameters$data[[1]]

# based on parameters
parameters %>% 
   filter(mean_tacos == 5, sd_tacos == 1) %>% 
   pull(data) %>% 
   .[[1]]
```

1) Unnest all datasets and plot

```{r}
parameters %>% 
   unnest() %>% 
   ggplot(aes(x = tacos_eaten, y = points_scored)) +
   geom_point() +
   facet_grid(mean_tacos ~ sd_tacos) +
   geom_smooth(method = "lm")
```

2) Fit a model to each dataset and store that in another column

```{r}
parameters <- 
   parameters %>% 
   mutate(
      model = data %>% map(~ lm(points_scored ~ tacos_eaten, data = .)),
      coef = model %>% map(broom::tidy)
   ) %>% 
   print()
```

Look at coefficients for each model

```{r}
parameters$coef
```



