---
title: "irt_local_dependence"
output: 
  github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

load packages

```{r, message=FALSE}
library(tidyverse)
library(mirt)
```

simulate data with a dependency. we'll have 10000 students and 10 items. all items will have difficulty 0 and discrimination 1. the first 9 items will be as usual. the order dependence comes from students who got the 9rd item correct get a 1-unit boost to their ability for the 10th item. 

```{r}
# simulate first 9 items
set.seed(1)
n_students <- 100000
dim <- 1
n_items <- 9

data <- 
    simdata(
        a = matrix(rep(1, n_items), nrow = n_items, ncol = dim),
        d = matrix(rep(0, n_items), ncol = 1),
        guess = rep(0, n_items),
        N = n_students,
        sigma = diag(dim),
        mu = 0,
        itemtype = rep("3PL", n_items),
        returnList = TRUE
    )

# clean up results
results <- 
    data$data %>% 
    as_tibble() %>% 
    janitor::clean_names() %>% 
    mutate(theta_1to9 = data$Theta[ , 1]) %>% 
    mutate(theta_10 = ifelse(item_9 == 1, theta_1to9 + 1, theta_1to9)) %>% 
    select(starts_with("theta"), everything())

# add in 10th item
set.seed(2)

results$item_10 <- 
    simdata(
        Theta = matrix(results$theta_10, nrow = n_students, ncol = dim),
        a = matrix(1, nrow = 1),
        d = matrix(0, ncol = 1),
        guess = 0,
        itemtype = "3PL",
        returnList = FALSE
    )[, 1]

# check it out
results %>% select(starts_with("theta"), item_9, item_10)
```

first, let's verify that a model fit just to the first 9 items is as expected

```{r}
model_first_9 <- mirt(results %>% select(item_1:item_9), 1, "2PL", verbose = FALSE)

# indeed, parameter recovery is good
coef(model_first_9, simplify = TRUE)$items %>% 
    as_tibble() %>% 
    select(disc = 1, easy = 2)

# indeed, we fail to reject goodness of fit
M2(model_first_9)
```

second, let's see that a model fit to all items has some quirks

```{r}
model_all <- mirt(results %>% select(item_1:item_10), 1, "2PL", verbose = FALSE)

# as makes sense, the discrimination of 9 and 10 are boosted and the 10th item is easier 
coef(model_all, simplify = TRUE)$items %>% 
    as_tibble() %>% 
    select(disc = 1, easy = 2)

# indeed, we reject goodness of fit
M2(model_all)
```
