---
title: 'R isn''t fast, but your R code could probably be a lot faster: an example'
author: "Ben Stenhaug"
output: 
  github_document:
    toc: true
    toc_depth: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

A friend of mine was recently complaining that R is slow. It's true that R isn't the fastest language, but it's also true that the gap between someone's slow R code and the optimal code in another language is usually mostly bridged by writing better R code.

# The Example

My friend gave me the following description of his problem:

> "One file is one line per student demographic data in the entire district.  the other is a transcript file that's one line per course taken by each student in the district over their entire time in the district.  So there are ~7500 students, but the transcript file has ~325,000 entries.  my goal is to make a dichotomous variable for each math (~850 individual classes) and ELA class (~50 individual classes) offered in the district, add them as columns to the demographic dataset and give a student a 1 or a 0 if they took each class."

He also sent me his current code:

```{r eval=FALSE}
demo.credits = read.csv("demo.credits.csv", header = TRUE)
transcript.demo = read.csv("transcript-demo.csv", header = TRUE)

math.courses <- unique(transcript.demo$coursename[transcript.demo$credittype=="Math"])
ela.courses <- unique(transcript.demo$coursename[transcript.demo$credittype=="English"])
courses <- unlist(list(math.courses,ela.courses))

classes <- matrix(NA,length(demo.credits$studentnumber),length(courses))
colnames(classes) <- courses

for (i in 1:length(demo.credits$studentnumber)){
	class.list <- transcript.demo$coursename[transcript.demo$studentnumber==demo.credits$studentnumber[i]]
	for (j in 1:length(courses)){
		classes[i,j] <- ifelse(sum(class.list==courses[j])>1,1,0)
	}
	cat("student", i, "completed, ", (length(demo.credits$studentnumber)-i), "students remaining\n")
	flush.console()
}

demo.credits.class <- cbind(demo.credits, classes)
write.csv(demo.credits.class, "demo-credits-class.csv")
```

# Speeding It Up

The first thing I notice when looking at his code is that he approached the problem — with nested loops and the necessary logic — in a way that works well for other languagues, but isn't the best for R. In general, R is best if you can write vectorized code as opposed to loops. Bonus points for keeping data in rectangular data frames and using the tools of the tidyverse. 

The second thing I notice is that I don't have access  to his data. So I make a very simple example:

```{r}
demo <- 
   tribble(
      ~studentnumber,  ~sex,
      1, "male",
      2, "female",
      3, "male"
   )

transcript <- 
   tribble(
      ~studentnumber, ~classname, ~credittype,
      1, "algebra", "math",
      1, "poetry", "english",
      2, "poetry", "english",
      3, "algebra", "math"
   )
```

Now I can work to solve the problem. My approach is to first make a single variable for class which is the credit type (math or english) followed by the class name (algebra for example). I think this makes things easier to read our results.

I then use spread which takes a column of keys (the classes) and values (I make an indicator of 1 because if the student is in the dataset, they took the course) and then spreads those two columns wide into a column for each unique value in the key column. 

Usually combinations not seen in the dataset are left as NA but in this case I fill with 0 to encode that the student did not take the course. Notice that I also make 

```{r}
transcript_wide <- 
   transcript %>% 
   mutate(
      class = paste0(credittype, "_", classname), 
      indicator = 1
   ) %>% 
   select(studentnumber, class, indicator) %>% 
   spread(class, indicator, fill = 0)
```

Now all we need to do is join with our demographics data, and we have what we're looking for!

```{r}
final <- 
   demo %>% 
   left_join(transcript_wide, by = "studentnumber")

final
```

