---
title: "PCA"
---

# Describe this doc
You can "Run All" (with the "Ctrl + Alt + R"" keyboard shortcut) to execute PCA and imputation

```{r}
load("ben.Rdata")

library(assertr)
library(tidyverse)
#library(MASS)
library(mice)
```

# Load one of the survey_wide data files:
#   anonymous, original, or original + unmatched

```{r}
#survey <- read_csv("data_created/survey_wide_with_outcome_anon.csv")
#survey <- read_csv("data_created/survey_wide_with_outcome1.csv")
survey <- read_csv("data_created/survey_wide_with_outcome.csv")

```

# Add variables

```{r}
survey <- survey %>% 
  mutate(STRatM_district = a13_district / a14_district,
         STRatM_CMO = a13_CMO / a14_CMO) %>% 
  mutate(STRatR_district = a15_district / a16_district,
         STRatR_CMO = a15_CMO / a16_CMO) %>% 
  mutate(STRatS8_district = a9_district / (b2_full_district + b2_part_district),
         STRatS8_CMO = a9_CMO / (b2_full_CMO + b2_part_CMO)) 


```

# Define helper functions

```{r}
pr_items <- function(data, items){
  data %>% 
    dplyr::select(one_of(items)) %>% 
    prcomp(center = TRUE, scale = TRUE) %>% 
    .$x %>% 
    .[ , 1]
}

one_pca <- function(items, name, just_cmo = "none"){
  items_district <- paste0(items, "_district")
  items_cmo <- paste0(items, "_CMO")
  
  names(survey)
  
  tmp1 <- survey %>% dplyr::select(c("Treatment_OBS", items_district)) %>% na.omit()
  tmp1 <- tmp1 %>% 
    mutate(type = "district") %>% 
    dplyr::select(Treatment_OBS, type, everything()) %>% 
    set_names(str_replace(names(.),"_district", ""))
  
  tmp2 <- survey %>% dplyr::select(c("Treatment_OBS", items_cmo)) %>% na.omit()
  tmp2 <- tmp2 %>% 
    mutate(type = "CMO") %>% 
    dplyr::select(Treatment_OBS, type, everything()) %>% 
    set_names(str_replace(names(.),"_CMO", ""))
  
  combined <- bind_rows(tmp1, tmp2)
  
  combined[[name]] <- pr_items(combined, items)
  
  combined %>% dplyr::select(-one_of(items))
}
```


# Run PCA

```{r}
# Helper function to expand by type
# TODO: there must be a standard way to do this...
expand_with_column <- function(df, colname, val1, val2) {
  df1 <- df %>%
    mutate(!!colname := val1)
  df2 <- df %>%
    mutate(!!colname := val2)
  bind_rows(df1, df2)
}

after_pca <- survey %>%
  dplyr::select(Treatment_OBS, contains("Treatment_CMOID_CMO"), math_outcome:SEReading_outcome, DistrictName_district) %>%
  expand_with_column("type", "CMO", "district") %>%
  left_join(one_pca(c("c7_a", "c7_b", "c7_c", "c7_d"), "StuSanct")) %>%
  left_join(one_pca(c("STRatM", "STRatR", "STRatS8"), "StuMon")) %>%
  left_join(one_pca(c("c5a_a", "c5a_b", "c5a_c", "c5a_f", "c5b_a", "c5b_b", "c5b_c", "c5b_f"), "TchrMon")) %>%
  left_join(one_pca(c("c5a_d", "c5a_e", "c5b_d", "c5b_e", "b20_b", "b20_c"), "TchrInst")) %>%
  left_join(one_pca(c("b3", "d2_b", "c11_a", "c11_b"), "TchrSanct")) %>%
  left_join(one_pca(c("c6_a", "c6_b", "c6_c"), "TchrColl")) %>%
  left_join(one_pca(c("b9_i", "b9_j", "b9_k", "b9_l", "b9_m"), "TchrScreen")) %>%
  #left_join(one_pca(c("c5b_a", "c5b_b", "c5b_c", "c5b_f"), "TchrMonExp")) %>%
  #left_join(one_pca(c("c5a_a", "c5a_b", "c5a_c", "c5a_f"), "TchrMonNew")) %>%
  left_join(one_pca(items = c("a17_d", "a17_e", "a17_f", "a17_g"), "FamCom"))

after_pca <- as.data.frame(after_pca) %>%
  dplyr::select(type, Treatment_OBS, Treatment_CMOID_CMO, everything()) %>%
  arrange(Treatment_OBS, type)
```

# Run PCA (Ben's version)

```{r, eval=FALSE}
# within one_pca the first argument is items for both, the second argument is the name of the column, and the third argument is items for just the cmo
#after_pca <- survey %>% dplyr::select(Treatment_OBS, KIPP, math_outcome:SEReading_outcome) %>%
after_pca <- survey %>% dplyr::select(Treatment_OBS, contains("Treatment_CMOID_CMO"), math_outcome:SEReading_outcome) %>%
  left_join(one_pca(c("c7_a", "c7_b", "c7_c", "c7_d"), "StuSanct")) %>% # works
  left_join(one_pca(c("STRatM", "STRatR", "STRatS8"), "StuMon")) %>% # works
  left_join(one_pca(c("c5a_a", "c5a_b", "c5a_c", "c5a_f", "c5b_a", "c5b_b", "c5b_c", "c5b_f"), "TchrMon")) %>% # works
  left_join(one_pca(c("c5a_d", "c5a_e", "c5b_d", "c5b_e", "b20_b", "b20_c"), "TchrInst")) %>% # works
  left_join(one_pca(c("b3", "d2_b", "c11_a", "c11_b"), "TchrSanct")) %>% # works
  left_join(one_pca(c("c6_a", "c6_b", "c6_c"), "TchrColl")) %>% # works
  left_join(one_pca(c("b9_i", "b9_j", "b9_k", "b9_l", "b9_m"), "TchrScreen")) %>% # works
  #left_join(one_pca(c("c5b_a", "c5b_b", "c5b_c", "c5b_f"), "TchrMonExp")) %>%
  #left_join(one_pca(c("c5a_a", "c5a_b", "c5a_c", "c5a_f"), "TchrMonNew")) %>%
  left_join(one_pca(items = c("a17_d", "a17_e", "a17_f", "a17_g"), name = "FamCom")) # works

after_pca <- as.data.frame(after_pca) %>%
  dplyr::select(type, Treatment_OBS, Treatment_CMOID_CMO, everything()) %>%
  arrange(Treatment_OBS, type)
```

# Reshape back to wide

```{r}
after_pca2 <- 
  after_pca %>% 
  gather(var, val, StuSanct:FamCom) %>% 
  mutate(newvar = paste0(var, "_", type)) %>% 
  dplyr::select(-type, -var) %>% 
  spread(newvar, val) %>% 
  dplyr::select(-contains("_NA")) 
```

# Try running MICE
This is where we'll actually use mice to run models

```{r}
# this runs mice (with m = 5 datasets produced)
nimp <- 5
imputed_data <- mice(after_pca2, m = nimp, maxit = 50, meth = 'pmm', seed = 500)
```

# Define MICE helper functions for models

```{r}
lmMICE <- function(input_model) {
  mi <- vector("list", nimp)
  for (i in 1:nimp) {
    data <- mice::complete(imputed_data, i)
    processed <- data %>% 
      left_join(survey %>% dplyr::select(Treatment_OBS, state = state_CMO, cmo_id = Treatment_CMOID_CMO,  
                                         district_name =  DistrictName_CMO, leaid = LEAID_CMO))
    ## Run least squares on each imputed dataset,
    ## and save results in a list vector
    mi[[i]] <- lm(input_model, data=processed)
  }

  ## Draw 1/nimp of the beta simulations from each run of least squares
  sims <- 10000
  simbetas <- NULL
  for (i in 1:nimp) {
    simbetas <- rbind(simbetas, MASS::mvrnorm(sims/nimp, coef(mi[[i]]), vcov(mi[[i]])))
  }

  pe <-apply(simbetas,2,mean)
  se <-apply(simbetas,2,sd)
  tstat <- pe/se
  pval <- 2*pt(-abs(tstat),  df.residual(mi[[1]]))
  lo95 <- apply(simbetas, 2, quantile, probs=0.025)
  up95 <- apply(simbetas, 2, quantile, probs=0.975)

  cbind(pe,se,pval,lo95,up95)
}

i <- 1 #TODO: why is this here?

```

